The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-31 script/en/chapters/numbertheory.tex
fixed various table to use the table environment and caption

If you allow numbers other than primes as factors, there are several ways of factorizing integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % should be \pageref{remFundTheoOfArithm}, but
% hyperref seems to be buggy
at the end of chapter \ref{primesinmath}.
}
for which numbers in the 
set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can
be used to test extremely large numbers to determine whether they are prime),
please refer to the article on \hyperlink{Kapitel_2}{prime numbers, 
chapter \ref{Label_Kapitel_2}} in this script.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\index{Modulus} \index{Divisibility}

@@

\begin{example}{:}\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$); but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem \ref{thm-zth-div} does not even apply to sums in one direction.
It is correct for sums only if the remainder is 0 and only in the following direction:
If a divisor divides both summands with no remainder, it also divides the sum with no remainder.
\end{example}

We can apply the above equivalence in theorem \ref{thm-zth-div} if we need a quick and easy method of determining whether 
large numbers are divisible by a certain number.

@@

From algebra theorems it follows that essential parts of the conventional
calculation rules are kept when we proceed to modular calculations over
a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a
division\footnote{%
\label{ftn-zth-divmodn}
When dividing modulo $m$\index{Division modulo $n$} we can one use
co-prime\index{Number!co-prime} numbers because because other numbers
have the same property as zero. See footnote
\ref{ftn-mod6} in chapter \ref{addmult}.
} is not a fraction but rather a whole number between $0$ and $m-1$.

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 = -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{NumberTheory_Appendix_C}{Appendix C of this chapter}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{\label{ftn-mod6}Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
For all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}

% --------------------------------------------------------------------------
\subsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsection*{Example of a multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

A counterexample shows the multiplication table for mod $6$ (since the modulus $m=6$ is not prime, not all elements from 
$\mathbb{Z}_6\setminus \{0\}$ are relatively prime to $6$):

\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
Multiplication table modulo $6$: \quad* &  1 & 2 & 3 & 4 & 5  \\
\hline 
1 &  1 & 2 & 3 & 4 & 5 \\  
2 &  2 & {\bf 4} & {\bf 0} & {\bf 2} & 4 \\
3 &  3 & {\bf 0} & {\bf 3} & {\bf 0} & 3 \\
4 &  4 & {\bf 2} & {\bf 0} & {\bf 4} & 2 \\
5 &  5 & 4 & 3 & 2 & 1 \\
\end{tabular}  
\end{center} 
\end{table}

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source
code to compute the tables using Sage.}
illustrate the properties of multiplicative inverses.

In the multiplication table mod $17$, the following was calculated for $i = 1, 2, \cdots, 18$:
\begin{itemize}
   \item[] $(5*i)/17 = a$ remainder $r$ and high-lighted $5*i \equiv 1$ (mod $17$),
   \item[] $(6*i)/17 = a$ remainder $r$ and high-lighted $6*i \equiv 1$ (mod $17$).
\end{itemize}
We need to {\bf find} the $i$ for which the product remainder $a*i$ modulo $17$
with $a=5$ or $a=6$ has the value $1$.

\subsection*{Table 1: Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Relatively prime} to the modulus $m=17$.
\enlargethispage{0.5cm}

\noindent {\bf The multiplicative inverse of $5$ (mod $17$) is $7$, while the inverse of $6$ (mod $17$) is $3$.}

\vskip +20 pt

\subsection*{Table 2: Multiplication table modulo $13$ (for $a=5$ and $a=6$)}\label{SrcArith1b}
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17  & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 2  & 7  & 12  & 4 & 9  & {\bf 1}  & 6  & 11 & 3  & 8  & 0 & 5  & 10  & 2   & 7   & 12 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 12  & 5  & 11  & 4  & 10  & 3  & 9  & 2  & 8  & {\bf 1}  & 7  & 0  & 6  & 12  & 5   & 11   & 4 \\
\hline 
\end{tabular}
\end{center}

@@

\noindent {\bf The multiplicative inverse of $5$ (mod $13$) is $8$, while the inverse of $6$ (mod $13$) is $11$.}

\vskip +20 pt

The following table contains an example, where the modulus $m$ and the number $a=6$ are {\em not} relatively prime.

\subsection*{Table 3: Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

\noindent {\bf The multiplicative inverse of $5$ (mod $12$) is $5$. The number $6$ has no inverse (mod $12$).}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsection{Fast calculation of high powers} \label{hohpot}\index{Power}
RSA encryption and decryption\footnote{See chapter \ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{NumberTheory_Appendix_A}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem \ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

If $J(m)$ cannot be determined\footnote{According to the first fundamental
theorem of number theory\index{Number theory!fundamental theorem} and
theorem \ref{thm-zth-phinum}, we can determine $J(m)$ by reducing $m$ to
prime factors\index{Prime factor!decomposition}.
}, it is difficult to calculate the $c$-th root. This forms the basis for the
security assumption used by the RSA encryption system (see chapter
\ref{rsabeweis} or chapter \ref{rsaverfahren}).

\end{itemize}
The time required for inverting addition and multiplication, on the other hand, 
is simply proportional to $\log m$ or $(\log m)^2$. Powers (for a number $x$ 
calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions
\index{One way function}
(compare chapters \ref{OneWayFunktion1} and \ref{OneWayFunktion2}).

@@

% --------------------------------------------------------------------------
\hypertarget{patternsandstructures}{}
\subsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

\begin{theorem}\label{thm-zth-phimultprime}\label{J_of_p1..pk}
If $n = p_1 * p_2 * \cdots * p_k$ where $p_1$ to $p_k$ are distinct prime numbers (i.e. no factor occurs more than once), 
then the following is true (as a generalization of theorem \ref{thm-zth-phipq}):
$$J(n) = (p_1 - 1)*(p_2 - 1)* \cdots *(p_k - 1).$$
\end{theorem}

>>>>

If you allow numbers other than primes as factors, there are several ways of factorizing integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % should be \pageref{remFundTheoOfArithm}, but
% hyperref seems to be buggy
at the end of chapter~\ref{primesinmath}.
}
for which numbers in the 
set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can
be used to test extremely large numbers to determine whether they are prime),
please refer to the article on \hyperlink{Kapitel_2}{prime numbers, 
chapter~\ref{Label_Kapitel_2}} in this script.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.
}
\index{Modulus} \index{Divisibility}

@@

\begin{example}{:}\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$); but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem~\ref{thm-zth-div} does not even apply to sums in one direction.
It is correct for sums only if the remainder is 0 and only in the following direction:
If a divisor divides both summands with no remainder, it also divides the sum with no remainder.
\end{example}

We can apply the above equivalence in theorem~\ref{thm-zth-div} if we need a quick and easy method of determining whether 
large numbers are divisible by a certain number.

@@

From algebra theorems it follows that essential parts of the conventional
calculation rules are kept when we proceed to modular calculations over
a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a
division\footnote{%
\label{ftn-zth-divmodn}
When dividing modulo $m$\index{Division modulo $n$} we can one use
co-prime\index{Number!co-prime} numbers because because other numbers
have the same property as zero. See footnote
\ref{ftn-mod6} in chapter~\ref{addmult}.
} is not a fraction but rather a whole number between $0$ and $m-1$.

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 = -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{NumberTheory_Appendix_C}{Appendix C of this chapter}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{\label{ftn-mod6}Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
For all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote~\ref{ftn-zth-divmodn}.} (table~\ref{addmod5})
for mod $5$ (table~\ref{mulmod5}) and mod $6$ (table~\ref{mulmod6}).
\end{itemize}

% --------------------------------------------------------------------------
\subsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
+ &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3
\end{tabular} 
\end{center} 
\caption{Addition table modulo 5}
\label{addmod5}
\end{table}

% --------------------------------------------------------------------------
\subsection*{Example of a multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
* & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2 \\
4 & 4 &    3    &    2    & 1
\end{tabular}
\end{center}
\caption{Multiplication table modulo 5}
\label{mulmod5}
\end{table}

@@

Table~\ref{mulmod6} shows as a counterexample the multiplication table for mod $6$ (since the modulus $m=6$ is not prime, not all elements from 
$\mathbb{Z}_6\setminus \{0\}$ are relatively prime to $6$).

\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
* &  1 & 2 & 3 & 4 & 5  \\
\hline 
1 &  1 & 2 & 3 & 4 & 5 \\  
2 &  2 & {\bf 4} & {\bf 0} & {\bf 2} & 4 \\
3 &  3 & {\bf 0} & {\bf 3} & {\bf 0} & 3 \\
4 &  4 & {\bf 2} & {\bf 0} & {\bf 4} & 2 \\
5 &  5 & 4 & 3 & 2 & 1
\end{tabular}  
\end{center} 
\caption{Multiplication table modulo $6$}
\label{mulmod6}
\end{table}

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem~\ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter~\ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source
code to compute the tables using Sage.}
illustrate the properties of multiplicative inverses.

In the multiplication table mod $17$ (table~\ref{mulmod17}), the following was calculated for $i = 1, 2, \cdots, 18$:
\begin{itemize}
   \item[] $(5*i)/17 = a$ remainder $r$ and high-lighted $5*i \equiv 1$ (mod $17$),
   \item[] $(6*i)/17 = a$ remainder $r$ and high-lighted $6*i \equiv 1$ (mod $17$).
\end{itemize}
We need to {\bf find} the $i$ for which the product remainder $a*i$ modulo $17$
with $a=5$ or $a=6$ has the value $1$.

\begin{table}[h]
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}\label{mulmod17}
\end{table}
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Relatively prime} to the modulus $m=17$.
\enlargethispage{0.5cm}

{\bf The multiplicative inverse of $5$ (mod $17$) is $7$, while the inverse of $6$ (mod $17$) is $3$.}

\begin{table}[h]
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17  & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 2  & 7  & 12  & 4 & 9  & {\bf 1}  & 6  & 11 & 3  & 8  & 0 & 5  & 10  & 2   & 7   & 12 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 12  & 5  & 11  & 4  & 10  & 3  & 9  & 2  & 8  & {\bf 1}  & 7  & 0  & 6  & 12  & 5   & 11   & 4 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $13$ (for $a=5$ and $a=6$)}\label{SrcArith1b}\label{mulmod13}
\end{table}

@@

{\bf The multiplicative inverse of $5$ (mod $13$) is $8$, while the inverse of $6$ (mod $13$) is $11$.}

The following table~\ref{mulmod12} contains an example, where the modulus $m$ and the number $a=6$ are {\em not} relatively prime.

\begin{table}[h]
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\label{mulmod12}
\end{table}
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

{\bf The multiplicative inverse of $5$ (mod $12$) is $5$. The number $6$ has no inverse (mod $12$).}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsection{Fast calculation of high powers} \label{hohpot}\index{Power}
RSA encryption and decryption\footnote{See chapter~\ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

The situation is similar if 
$b$ is the unknown variable and we know the values of $a$ and $c$: \\
If we know the Euler function \index{Euler!(phi) function} of $m, J(m)$, then we can easily\footnote{See \hyperlink{NumberTheory_Appendix_A}{Appendix A of this chapter}:
the greatest common divisor (gcd) of whole numbers.}\index{Gcd} calculate $d$ with $c*d \equiv 1 {\rm ~(mod~} J(m))$ and 
use theorem~\ref{thm-zth-pot} to obtain:
$$
   a^d \equiv (b^c)^d \equiv b^{c*d} \equiv b^{c*d~(mod~J(m))} \equiv b^1 \equiv b {\rm ~(mod~} m)
$$
the {\em $c$-th root} $b$ of $a$. \par

If $J(m)$ cannot be determined\footnote{According to the first fundamental
theorem of number theory\index{Number theory!fundamental theorem} and
theorem~\ref{thm-zth-phinum}, we can determine $J(m)$ by reducing $m$ to
prime factors\index{Prime factor!decomposition}.
}, it is difficult to calculate the $c$-th root. This forms the basis for the
security assumption used by the RSA encryption system (see chapter
\ref{rsabeweis} or chapter~\ref{rsaverfahren}).

\end{itemize}
The time required for inverting addition and multiplication, on the other hand, 
is simply proportional to $\log m$ or $(\log m)^2$. Powers (for a number $x$ 
calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions
\index{One way function}
(compare chapters~\ref{OneWayFunktion1} and \ref{OneWayFunktion2}).

@@

% --------------------------------------------------------------------------
\hypertarget{patternsandstructures}{}
\subsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter~\ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

\begin{theorem}\label{thm-zth-phimultprime}\label{J_of_p1..pk}
If $n = p_1 * p_2 * \cdots * p_k$ where $p_1$ to $p_k$ are distinct prime numbers (i.e. no factor occurs more than once), 
then the following is true (as a generalization of theorem~\ref{thm-zth-phipq}):
$$J(n) = (p_1 - 1)*(p_2 - 1)* \cdots *(p_k - 1).$$
\end{theorem}
